---
title: Plotting prior distributions with ggplot2
author: Stefano Coretta
date: '2019-06-17'

categories:
  - Methods
  - Regression Models
tags:
  - data viz
  - rstats
  - tidy data
  - statistics
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = here::here())
```

The choice of priors is a fundamental step of the Bayesian inference process. @vasishth2018 recommend plotting the chosen priors to see if they are reasonable.

In this post I will show how to easily plot prior distributions in [ggplot2](https://ggplot2.tidyverse.org) (which is part of the [tidyverse](https://www.tidyverse.org)).

Let's load the tidyverse first.

```{r tidy}
library(tidyverse)
theme_set(theme_minimal()) # I just like this theme :)
```

## Plotting your priors

Let's start with a simple normal prior with $\mu$ = 0 and *sd* = 1.

The plot is initialised with an empty call to `ggplot()`.
As aesthetics, you only need to specify the range of *x* values in `aes()`.
Here, we use `c(-4, 4)`, meaning that the *x*-axis of this plot will have these limits.
For a normal distribution, it is useful to set the limits as the mean ± 4 times the standard deviation (this ensures all the distribution is shown).

The function `ggplot2::stat_function()` allows us to specify a distribution family with the `fun` argument.
This arguments takes the density function (the R functions of the form *dxxx*) of the chosen distribution family, so for the normal (Gaussian) distribution we use `dnorm()`.
The argument `n` specifies the number of points along which to calculate the distribution (here `101`), while `args` takes a list with the parameters of the distribution (here the mean `0` and standard deviation `1`).

```{r normal}
ggplot(data = tibble(x = -4:4), aes(x)) +
  stat_function(fun = dnorm, n = 101, args = list(1)) +
  labs(title = "Normal (Gaussian) distribution")
```

A beta prior will be bounded between 0 and 1, so we can specify that in `aes()`.
The beta distribution has two arguments, `shape1` and `shape2` (here `2` and `5`).

```{r beta}
ggplot(data = tibble(x = 0:1), aes(x)) +
  stat_function(fun = dbeta, n = 101, args = list(2, 5)) +
  labs(title = "Beta distribution")
```

Another common distribution is the Cauchy.

```{r cauchy}
ggplot(data = tibble(x = -10:10), aes(x)) +
  stat_function(fun = dcauchy, n = 201, args = list(-2, 1)) +
  labs(title = "Cauchy distribution")
```

The Poisson distribution can be plotted by changing the type of `geom` and using an `n` that creates only integers.

```{r poisson}
# the range 0:20 includes 21 integers, so n = 21
ggplot(data = tibble(x = 0:20), aes(x)) +
  stat_function(fun = dpois, n = 21, args = list(4), geom = "point") +
  labs(title = "Poisson distribution")
```

Of course any family with a corresponding *dxxx* function can be plotted (see `?Distributions` and package-provided families).

## References


