---
title: How to globally set colour scales in ggplot2
author: Stefano Coretta
date: '2020-12-22'
slug: []
categories:
  - Methods
tags:
  - data viz
  - ggplot2
  - colours
  - rmd
  - rmarkdown
lastmod: '2020-12-28'
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
output:
  blogdown::html_page:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

After the post on using custom typefaces in ggplot2 ([here](../custom-fonts-ggplot2)), this time I'll briefly discuss how to set colour scales in ggplot2 globally in an `.Rmd` file.

The perk of setting scales globally is that you can set the colours once at the beginning of the `.Rmd` file and all the plots in the file will adhere to the specified scales without the need to repeat code.

# Set ggplot2 options

The most robust way of specifying global colour scales for ggplot2 plots is to set the appropriate ggplot2 options with `options()` at the beginning of the `.Rmd` file.
I generally do so in the `setup` chunk, where I also load the tidyverse/ggplot2 with `library(tidyverse)`.

The relevant options are:

- `ggplot2.discrete.colour`: sets the colour scale for discrete aesthetics.
- `ggplot2.discrete.fill`: sets the fill scale for discrete aesthetics.
- `ggplot2.continuous.colour`: sets the colour scale for continuous aesthetics.
- `ggplot2.continuous.fill`: sets the fill scale for continuous aesthetics.

All these options take a character vector of colour names/codes or a function that returns such a vector (for the full documentation see [here](https://ggplot2.tidyverse.org/reference/scale_colour_discrete.html?q=discrete%20colour%20scales) and [here](https://ggplot2.tidyverse.org/reference/scale_colour_continuous.html)).

For example:

```{r set, eval=FALSE}
# names
options(ggplot2.discrete.fill = c("red", "blue", "green", "purple", "yellow"))

# hex codes
options(ggplot2.discrete.fill = c("#264653", "#2a9d8f", "#a8dadc", "#457b9d", "#e76f51"))

# function
options(ggplot2.discrete.fill = RColorBrewer::brewer.pal(12, "Set3"))
```

Note that the colours specified with `ggplot2.discrete.*` will be used only if there are as many colours in the vector/function output as the number of data levels, or more.
When the number of level exceeds the number of colours, ggplot2 falls back to `scale_colour_hue()/scale_fill_hue()` to construct the scale.

It is possible to define multiple palettes each with a different number of colours as a list of colour names/codes vectors, and ggplot2 will pick the palette with the minimum number of colours that exceeds the number of data levels.
See \@ref(interp) and the third example [here](https://ggplot2.tidyverse.org/reference/scale_colour_discrete.html).

# Colours in action

```{r tidy, message=FALSE}
library(tidyverse) # or library(ggplot2)
```

## Colour name

```{r name}
options(ggplot2.discrete.fill = c("red", "blue", "green", "purple", "yellow"))

mpg %>%
  filter(!(class %in% c("subcompact", "suv"))) %>%
  ggplot(aes(class, fill = class)) +
  geom_bar()
```

## Hex value

```{r hex}
options(ggplot2.discrete.fill = c("#264653", "#2a9d8f", "#a8dadc", "#457b9d", "#e76f51"))

mpg %>%
  filter(!(class %in% c("subcompact", "suv"))) %>%
  ggplot(aes(class, fill = class)) +
  geom_bar()
```

## Set3

```{r set3}
options(ggplot2.discrete.fill = RColorBrewer::brewer.pal(12, "Set3"))

mpg %>%
  filter(!(class %in% c("subcompact", "suv"))) %>%
  ggplot(aes(class, fill = class)) +
  geom_bar()
```

## Palette interpolation {#interp}

```{r ramp}
default_palettes <- list(
  # palette with 5 colours
  wesanderson::wes_palette("Darjeeling1"),
  # same palette interpolated to 8 colours
  grDevices::colorRampPalette(wesanderson::wes_palette("Darjeeling1"), alpha = TRUE)(8)
)

options(ggplot2.discrete.fill = default_palettes)

# 5 levels
ggplot(mpg, aes(class, fill = fl)) + geom_bar()

# 7 levels
ggplot(mpg, aes(class, fill = class)) + geom_bar()
```
